
---
title: "Predicting GTEx by a TF-Only Model Learned from TCGA"
author: "Amir Asiaee"
date: '`r Sys.Date()`'
output:
  html_document:
    highlight: kate
    theme: yeti
    toc: yes
---

```{r globopt, echo=FALSE}
knitr::opts_chunk$set(fig.width=12) 
```

```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Getting Started
Here the goal is to ignore the miRs and also Train/Test. We assume that the correct number of parameters is 15 for clusters and we learn 15 clusters using only TFs. And we do not do train/test split, we use the whole TCGA data (for which we already know the thresher's output for TFs) once to learn 15 clusters and for each one of them a linear model. Then we test that linear model on GTEx.

```{r globals}
rm(list=ls())
source("00-paths.R")
ls()
```


# Load the Relevant Data
```{r bigModel, warning = FALSE}

load(file.path(paths$scratch, paste('medoids_tf_all.Rda', sep = '')))
dim(tfMedoid)
load(file.path(paths$scratch, 'mrnaRestdata.Rda'))
dim(mRNARest)
```


# Regression

```{r helpers, warning = FALSE, echo=FALSE}
library(ClassComparison)
library(oompaBase)

computeCor <- function(res, Y, ns=10^4){
  Yhat <- Y - res
  M <- nrow(Y)
  Ours <- sapply(1:M, function(i) cor(Y[i,], Yhat[i,]))
  
  Yind <- sample(1:M, ns, replace = T)
  YhatInd <- sample(1:M, ns, replace = T)
  ind <- cbind(Yind, YhatInd)
  ind <- ind[Yind != YhatInd, ]
  Null <- sapply(1:nrow(ind), function(j) cor(Y[ind[j,1],], Yhat[ind[j,2],]))
  z <- quantile(Null, .95, na.rm = T)
  acc <- sum(Ours > z, na.rm = T) / M
  return(list(ours=Ours, null=Null, acc=acc))
}


computeR2 <- function(res, Y){
  M <- ncol(Y)
  dmean <- matrixMean(Y)
  dvar <- as.vector(matrixVar(Y, dmean))
  ssres <- apply(res ^ 2, 1, sum)
  r2 <- rep(0, M)
  # varInd <- dvar != 0
  # r2[varInd] <- 1 - ssres[varInd] / (dvar[varInd] * (M - 1))
  r2 <- 1 - ssres / (dvar * (M - 1))
  return(r2)
}

```

```{r , warning = FALSE}
nc <- 15 #20
f <- file.path(paths$scratch, paste('tf_final_model_', nc, '.Rda', sep = ''))
if (!file.exists(f)) {
  Yavg <- apply(mRNARest, 1, mean)
  Y <- sweep(mRNARest, 1, Yavg)
  X <- cbind(tfMedoid)
  
  coeffMatList <- list()
  N <- nrow(X); M <- nrow(Y); P <- ncol(X);
  res <- matrix(0, nrow = M, ncol = N)
  
  cat(paste("\n  Learning a model with ", nc, " clusters:"))
  cat("\n    1) Clustering")
  myClust <- kmeans(X, nc, iter.max = 100, nstart = 100)
  assignment <- myClust$cluster
  
  cat("\n    2) Regression")
  cat("\n      Fitting OLS to clusters 1")
  
  for (cluster in 1:nc) {
    if(cluster %% 5 == 0) cat(paste(cluster)) else cat('.')
    coeffMat <- matrix(0, nrow = P + 1, ncol = M, 
                       dimnames = list(c('(Intercept)',
                                         gsub('\\-', '.', colnames(X))), 
                                         rownames(Y)))
    
    S <- (assignment == cluster) #Cluster Selector
    print(paste("Size of cluster=", sum(S)))
    varCov <- apply(X[S,], 2, var) != 0
    Xs <- X[S, varCov]
    Ys <- Y[, S]
    
    if (nrow(Xs) < ncol(Xs)) {
      print("Not enough samples, using the means.")
      coeffMat[1,] <- apply(Ys, 1, mean)#YcAvg 
      res[, S] <- sweep(Ys, 1, coeffMat[1,])#mean0Yc 
    } else{
      mlm <- MultiLinearModel(Y ~ ., Xs, Ys)
      coeffMat[rownames(mlm@coefficients),] <- mlm@coefficients
      res[, S] <- Ys - t(mlm@predictions)
    }
    coeffMatList[[cluster]] <- coeffMat
  } 
  
  r2 <- computeR2(res, Y)
  cat(paste("\n      Average R2:", round(mean(r2, na.rm =T), 2)))
  rmse <- sqrt((1/N) * apply(res ^ 2, 1, sum))
  cat(paste("\n      Averge RMSE:", round(mean(rmse),2)))
  corr <- computeCor(res, Y)
  cat(paste("\n      Averge Cor:", round(mean(corr$acc),2)))
  
  model <- list(coeffMatList=coeffMatList, nc=nc, 
                  res=res, r2=r2, rmse=rmse, corr=corr,
                  centers=myClust$centers, alpha=Yavg)
  save(model, file = f)
} else {
  load(f)
}

hist(model$r2, breaks=123, main="R2", xlab=paste("Average R2 =", round(mean(model$r2, na.rm =T), 2)))
print(paste("\n      Average R2:", round(mean(model$r2, na.rm =T), 2)))
print(paste("\n      Average Corr:", round(mean(model$corr$acc, na.rm =T), 2)))

rm(coeffMat, coeffMatList, mlm, res, r2, rmse, corr, Yavg, myClust, Y, Ys, X, Xs)
```

# Testing on GTEx

## Find Matching Medoids in GTEx

Below, we load the separately normalized mRNA and miRNA data and then we select the TF and miR medoids from them:
```{r load gtext, warning = FALSE}
load(file.path(paths$scratch, "gtexMrnaRest2.Rda"))
load(file.path(paths$scratch, "gtexTFMed.Rda"))
dim(gtexMrnaRest)
dim(gtexTFMed)
```


### Testing

Load the learned model from TCGA:
```{r model load, warning = FALSE}
load(file.path(paths$scratch, paste('tf_final_model_', nc, '.Rda', sep = '')))
f <- file.path(paths$scratch, paste('tf_gtex_test_', model$nc, '.Rda', sep = ''))

if(!file.exists(f)){
  Y <- sweep(gtexMrnaRest, 1, model$alpha[rownames(gtexMrnaRest)])
  X <- cbind(gtexTFMed)
  
  centers <- model$centers
  cat(paste("\n  Testing with ", nrow(centers), " clusters:"))
  M <- nrow(Y); N <- ncol(Y); #nrow(X) == N
  D <- matrix(0, nrow = N, ncol = nrow(centers))
  for(i in 1:nrow(centers)){
    center <- centers[i, ]
    D[,i] <- apply(X, 1, function(x) sum((x - center)^2))
  }
  C <- apply(D, 1, which.min)
  
  res <- matrix(0, nrow = M, ncol = N)
  aX <- cbind(rep(1, nrow(X)), X) #augmented X
  for(i in 1:nrow(centers)){
    Beta <- model$coeffMatList[[i]]
    res[, C==i] <- Y[,C==i] - t(aX[C==i, ] %*% Beta[,rownames(gtexMrnaRest)])
  }
  rmse <- sqrt((1/N) * apply(res ^ 2, 1, sum))
  cat(paste("\n    Averge RMSE:", round(mean(rmse),2)))
  corr <- computeCor(res, Y)
  cat(paste("\n    Averge Cor:", round((corr$acc),2)))
  gtexResults <- list(res=res, rmse=rmse, corr=corr)
  save(gtexResults, file = f)
} else{
  load(f)
}

cat(paste("\n    Averge RMSE:", round(mean(gtexResults$rmse),2)))
cat(paste("\n    Averge Cor:", round((gtexResults$corr$acc),2)))


```
Not that bad! What if we only focus on genes that they predict? 

```{r subset prediction, warning = FALSE}
l1k <- read.csv(file.path(paths$raw, "12k.csv"))
l1k <- l1k[2:nrow(l1k),]

length(l1k$Gene.Symbol) - sum(l1k$Gene.Symbol %in% rownames(gtexMrnaRest))
toPredict <- l1k$Gene.Symbol[l1k$Gene.Symbol %in% rownames(gtexMrnaRest)]
```

```{r , warning = FALSE}
load(file.path(paths$scratch, paste('tf_final_model_', 15, '.Rda', sep = '')))
f <- file.path(paths$scratch, paste('tf_gtex_test_subset', model$nc, '.Rda', sep = ''))

gsubset <- gtexMrnaRest[toPredict, ]
if(!file.exists(f)){
  Y <- sweep(gsubset, 1, model$alpha[rownames(gsubset)])
  X <- cbind(gtexTFMed)
  
  centers <- model$centers
  cat(paste("\n  Testing with ", nrow(centers), " clusters:"))
  M <- nrow(Y); N <- ncol(Y); #nrow(X) == N
  D <- matrix(0, nrow = N, ncol = nrow(centers))
  for(i in 1:nrow(centers)){
    center <- centers[i, ]
    D[,i] <- apply(X, 1, function(x) sum((x - center)^2))
  }
  C <- apply(D, 1, which.min)
  
  res <- matrix(0, nrow = M, ncol = N)
  aX <- cbind(rep(1, nrow(X)), X) #augmented X
  for(i in 1:nrow(centers)){
    Beta <- model$coeffMatList[[i]]
    res[, C==i] <- Y[,C==i] - t(aX[C==i, ] %*% Beta[,rownames(gsubset)])
  }
  rmse <- sqrt((1/N) * apply(res ^ 2, 1, sum))
  cat(paste("\n    Averge RMSE:", round(mean(rmse),2)))
  corr <- computeCor(res, Y)
  cat(paste("\n    Averge Cor:", round((corr$acc),2)))
  gtexResults <- list(res=res, rmse=rmse, corr=corr)
  save(gtexResults, file = f)
} else{
  load(f)
}

cat(paste("\n    Averge RMSE:", round(mean(gtexResults$rmse),2)))
cat(paste("\n    Averge Cor:", round((gtexResults$corr$acc),2)))
```
# Appendix

This analysis was performed using the following R packages.
```{r si}
sessionInfo()

```



