
---
title: "Generating Figures for the Paper"
author: "Amir Asiaee"
date: '`r Sys.Date()`'
output:
  html_document:
    highlight: kate
    theme: yeti
    toc: yes
---

```{r globopt, echo=FALSE}
knitr::opts_chunk$set(fig.width=12) 
```

```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Getting Started
Getting started!
```{r globals}
rm(list=ls())
source("00-paths.R")
ls()
rm(tcgapaths, tfacpaths)

load(file.path(paths$clean, "matchedData.Rda"))
# Removing GBM
rm(miRdata, mRNAdata)
length(cancerType)
gbmIndex <- which(cancerType == 'GBM')
cancerType <- cancerType[-gbmIndex]
cancerType <- droplevels(cancerType)
cohortColors <- cohortColors[-gbmIndex]
sampleType <- sampleType[-gbmIndex]

load(file.path(paths$scratch, paste('medoids_tf_all.Rda', sep = '')))
load(file.path(paths$scratch, paste('medoids_mir_all.Rda', sep = '')))
X1 <- as.data.frame(cbind(tfMedoid, mirMedoid))
X <- data.frame(X1, ctype = cancerType)
```


# Feature Selection
```{r all thresher, warning = FALSE, echo=FALSE}
cancerMean <- aggregate(. ~ ctype, X, mean)
cancerCenters <- as.matrix(cancerMean[, -1])
row.names(cancerCenters) <- as.character(cancerMean[,1])

library(beanplot)
baselev <- levels(cancerType)

outputDir <- file.path(paths$clean, 'figures') 
if (!dir.exists(outputDir)) 
  dir.create(outputDir)

for (k in 1:ncol(X1)) {
  scaler <- 1024
  ot <- order(cancerMean[,k+1])
  tempType <- factor(cancerType, levels=baselev[ot])
  png(file=file.path(outputDir, paste("component", k, ".png", sep='')), 
    pointsize=36, width=5*scaler, height=1.5*scaler)
  # beanplot(X1[,k] ~ tempType, bw="ucv", col=as.list(cohortColors[ot[]]), 
           # what=c(1,1,1,0), log = "", main=paste("Component", k))
  boxplot(X1[,k] ~ tempType, col=cohortColors[ot[]], main=paste("Component", k))
  dev.off()
}

```

# Classification

```{r gbm removal, warning = FALSE, echo=FALSE}
library(viridis)
library(e1071)
library(caret)
library(ComplexHeatmap)

f <- file.path(paths$scratch, 'cv-SVM-noGBM.rda')

if(file.exists(f)){
  load(f)
} else {
  nfolds <- 10
  nTrain <- nrow(X)
  medSvmList <- pmedSvmList <- cfmedSvmList <- list()
  foldsId <- sample(rep(1:nfolds, length.out = nTrain), replace = F)
  cfmedSvmMean <- matrix(0, nrow=31, ncol=31)
  for (k in 1:nfolds) {
    print(paste("Fold", k, "started."))
    testId <- which(foldsId == k)
    train <- X[-testId, ]; 
    test <- X[testId, ];
    
    medSvm <- svm(ctype ~ ., train)
    pmedSvm <- predict(medSvm, test[,-ncol(test)])
    cfmedSvm <- confusionMatrix(pmedSvm, test$ctype)
    print(cfmedSvm$overall)
  
    medSvmList[[k]] <- medSvm
    pmedSvmList[[k]] <- pmedSvm
    cfmedSvmList[[k]] <- cfmedSvm
    temp <- as.data.frame.matrix(cfmedSvm$table)
    cfmedSvmMean <- cfmedSvmMean + as.matrix(temp)
  }
  cfmedSvmMean <- cfmedSvmMean / nfolds
  cat("Average Accuracy:", mean(sapply(cfmedSvmList, function(x) x$overall['Accuracy'])))
  save(cfmedSvmMean, medSvmList, pmedSvmList, cfmedSvmList, file = f)
}

res <- 300
# png(file=file.path(paths$figures, 'classification.png'), 
           # width=16*res, height=9*res, res=res, bg="white")
tempM <- cfmedSvmMean
diag(tempM) <- NA
heatmap(as.table(tempM), Rowv=NA, Colv = NA, col=viridis(256), main = "Average Confusion Matrix for 10-fold Cross Validation", scale = 'none')
# dev.off()
  
cat("Average Accuracy:", mean(sapply(cfmedSvmList, function(x) x$overall['Accuracy'])))
rm(cfmedSvm,cfmedSvmList, cfmedSvmMean, medSvm, medSvmList, pmedSvm, pmedSvmList, train, foldsId, k, nfolds, nTrain, testId)
rm(mirMedoid, tfMedoid, temp)
```


```{r tsne, warning = FALSE, echo=FALSE}
library(Rtsne)
f <- file.path(paths$scratch, "tsneData-joint-truth.rda")
if (file.exists(f)) {
  load(f)
} else {
  set.seed(97531)
  cancers <- Rtsne(X1, initial_dims = 60)
  colvec <- cohortColors[as.character(cancerType)]
  sym <- c(16, 1, 8)[as.numeric(sampleType)]
  xcent <- tapply(cancers$Y[,1], list(cancerType), median)
  ycent <- tapply(cancers$Y[,2], list(cancerType), median)
  save(cancers, colvec, sym, xcent, ycent, file = f)
}

medSvm <- svm(ctype ~ ., X)
pmedSvm <- predict(medSvm, X)
cfmedSvm <- confusionMatrix(pmedSvm, X$ctype)
print(cfmedSvm$overall)
colvec1 <- cohortColors[as.character(pmedSvm)]
 
# res <- 300
# png(file=file.path(paths$figures, 'classification-tsne.png'), 
#            width=16*res, height=9*res, res=res, bg="white")

par(mfrow=c(1,2))

plot(cancers$Y, pch=sym, col=colvec, xlab="T1", ylab="T2", cex=0.8, main="Ground Truth")
text(xcent, ycent, levels(cancerType), col='black', cex=1.5)
 
library(viridis)
colvec2 <- rep(viridis(5)[3], length(colvec1))
colvec2[colvec != colvec1] <- viridis(5)[5]
plot(cancers$Y, pch=sym, col=colvec2, xlab="T1", ylab="T2", cex=0.8, main="Classification")
rect(par("usr")[1], par("usr")[3],
     par("usr")[2], par("usr")[4],
     col = "lavender")
par(new = TRUE)
plot(cancers$Y, pch=sym, col=colvec2, xlab="T1", ylab="T2", cex=0.8, main="Classification")
text(xcent, ycent, levels(cancerType), col='black', cex=1.5)

# dev.off()

par(mfrow=c(1,1))

    
rm(f)
```

# Regression

## Tissue-Agnostic & Aware - No Train/Test Split

```{r , warning = FALSE, echo=FALSE}
load(file.path(paths$scratch, "jointMedoidR2.Rda"))
allR2 <- data.frame(agnostic = r2s)
load(file.path(paths$scratch, "tissueMedoidR2.Rda"))
allR2 <- data.frame(allR2, aware = r2s)
rm(r2s)
```

```{r determine k by bic, warning = FALSE, echo=FALSE}
# Run the EM 10 times for the range of the K=1...5 components:
# Since we don't have observations in the whole range the mle estimate will not work. So we change it to mme:
valTrsh <- 1E-5

f <- file.path(paths$scratch, 'betaModels-agnostic-aware.Rda')
if(file.exists(f)){
  load(f)
} else {
  library(fitdistrplus)
  library(BetaModels)
  set.seed(1234)
  betaModels <- list()
  K <- 4
  B <- 10
  for (expR2 in colnames(allR2)) {
    tempData <- allR2[,expR2]
    n <- length(tempData)
    tempData[tempData < valTrsh] <- valTrsh
    tempData[tempData > 1 - valTrsh] <- 1 - valTrsh
    betaModels[[expR2]] <- list()
    for (k in 1:K) {
      tempList <- list()
      for (b in 1:B) {
        print(paste(expR2, ", k = ", k, ", b = ", b))
        index <- sample(1:n, n, replace = T)
        if (k == 1) {
          output <- fitdist(tempData[index], "beta", "mle")
          params <- matrix(c(1, output$estimate), ncol = 1)
        } else {
          output <- BetaMixture(tempData[index], K = k, forever = 50, epsilon = 0.02)
          params <- rbind(output@phi, output@mle)
        }
        rownames(params) <- c("pi", "alpha", "beta")
        tempList[[length(tempList)+1]] <- params
      }
      betaModels[[expR2]][[k]] <- tempList
    }
  }
  save(betaModels, file = f)
}
```

```{r bicCompute, warning = FALSE, echo=FALSE}
# Find the best k
#We need to compute the BIC score for each k and average that over the ten runs that we did. 

betaMixtureLL <- function(dataVec, params){
  likelihood <- rep(0, length(dataVec));
  for(i in 1:ncol(params)){
    likelihood <- likelihood + params[1, i] * dbeta(dataVec, params[2,i], params[3, i]);
  }
  return (sum(log(likelihood)));
}

betaMixtureBic <- function(dataVec, params){
  k <- ncol(params);
  n <- length(dataVec);
  loglikelihood <- betaMixtureLL(dataVec, params);
  return (-2 * loglikelihood + (3*k - 1) * log(n));
}
```

```{r plotBic, warning = FALSE, echo=FALSE}
#First preparing ggplot2 library:
library(ggplot2)
library(grid)
library(gridExtra)
theme_set(theme_minimal() + 
            theme(axis.line = element_line(colour = "black"),
                  panel.grid.major = element_blank(),
                  panel.grid.minor = element_blank(),
                  panel.border = element_blank(),
                  panel.background = element_blank(),
                  title = element_text(size = 16)));

plotBic <- function(dfData, dfSlope, tempData, expR2, title){
  x <- seq(0, 1, len = length(tempData))
  
  p0 <- ggplot(as.data.frame(tempData), aes(tempData)) + 
    geom_histogram(aes(y=..density..), color = "black", fill = "white", binwidth = .01) +
    xlab(expression(R^2)) + ylab('Density') + xlim(c(0,1))
  
  p1 <- ggplot(dfData, aes(x = x, y = y)) +
    geom_point()+ geom_line() +
    xlab("k: Number of Components") + ylab("BIC Score") + 
    geom_errorbar(aes(ymin = y - dy, ymax = y + dy), width=.05) + 
    xlim(c(1,K));

  p2 <- ggplot(dfSlope, aes(x = x, y = y)) +
    geom_point()+ geom_line() +
    xlab("k: Number of Components") + ylab("Derivative of BIC Score") + 
    geom_errorbar(aes(ymin = y - dy, ymax = y + dy), width=.05) + 
    xlim(c(1,K-1));

  grid.newpage()
  pushViewport(viewport(layout = grid.layout(
        nrow = 2,
        ncol = 3,
        heights = c(0.1, 0.9))))
  grid.text(title, vp = viewport(layout.pos.row = 1, layout.pos.col = 1:3))
  print(p0, vp = viewport(layout.pos.row = 2, layout.pos.col = 1))
  print(p1, vp = viewport(layout.pos.row = 2, layout.pos.col = 2))
  print(p2, vp = viewport(layout.pos.row = 2, layout.pos.col = 3))
}
```


```{r bic-k, warning=FALSE, echo=FALSE}
# Find the best k by taking the derivative:
K <- 4
B <- 10
bicDerivative <- bic <- vector("list", length(colnames((allR2))))
names(bicDerivative) <- names(bic) <- colnames((allR2))
titles <- c("Tissue-Agnostic Model", "Tissue-Aware Model")
for (i in 1:ncol(allR2)) {
  expR2 <- colnames(allR2)[i]
  title <- titles[i]
  tempData <- allR2[,expR2]
  tempData[tempData < valTrsh] <- valTrsh
  tempData[tempData > 1 - valTrsh] <- 1 - valTrsh
  
  bicDerivative[[expR2]] <- bic[[expR2]] <- matrix(0, nrow = B, ncol = K);
  # For plotting
  dfData <- data.frame(matrix(nrow = K, ncol = 3))
  dfSlope <- data.frame(matrix(rep(0, K*3), nrow = K, ncol = 3))
  colnames(dfData) <- colnames(dfSlope) <- c('x', 'y', 'dy')
  for (k in 1:K) {
    print(paste(expR2, ", k = ", k))
    for (b in 1:B) {
      params <- betaModels[[expR2]][[k]][[b]]
      bic[[expR2]][b, k] <- betaMixtureBic(tempData, params)
      if(k > 1)
        bicDerivative[[expR2]][b, k-1] <- bic[[expR2]][b, k] - bic[[expR2]][b, k-1]
    }
    dfData[k,] <- c(k,
                    mean(bic[[expR2]][,k]),
                    sd(bic[[expR2]][,k]))
    if(k > 1){ 
      dfSlope[k-1,] <- c(k-1,
                       mean(bicDerivative[[expR2]][,k-1]),
                       sd(bicDerivative[[expR2]][,k-1]))
    }
  }
  plotBic(dfData, dfSlope, tempData, expR2, title)
}
```


```{r bestk, warning = FALSE, , echo=FALSE}
#Fitting the best K once with more precision:
library(fitdistrplus)
bestK <- c(2, 2) #manually entered 
bestBetaParameters <- bestBetaModels <- list()


bestModelsFileName <- file.path(paths$scratch, 'bestBetaModels-final-ag-aw.Rda')
bestParamsFileName <- file.path(paths$scratch, 'bestBetaParams-final-ag-aw.Rda')
if (!file.exists(bestParamsFileName)) {
  for (i in 1:ncol(allR2)) {
    expR2 <- colnames(allR2)[i]
    tempData <- allR2[, expR2]
    tempData[tempData < 1E-10] <- 1E-10
    tempData[tempData > 1 - 1E-10] <- 1 - 1E-10
    myK <- bestK[i]
    
    if (myK == 1) {
      output <- fitdist(tempData, "beta", "mle")
      params <- matrix(c(1, output$estimate), ncol = 1)
    } else {
      output <-
        BetaMixture(tempData,
                    K = myK,
                    forever = 100,
                    epsilon = 0.001)
      params <- rbind(output@phi, output@mle)
    }
    bestBetaModels[[expR2]] <- output
    rownames(params) <- c("pi", "alpha", "beta")
    bestBetaParameters[[expR2]] <- params
  }
  save(bestBetaParameters, file = bestParamsFileName)
  save(bestBetaModels, file = bestModelsFileName)
  
} else {
  load(bestParamsFileName)
  load(bestModelsFileName)
}
```


```{r , warning = FALSE, echo=FALSE}

library(ggplot2)
library(Polychrome)
myColors <- light.colors(5)
titles <-
  c("Transcriptome Prediction with 56 Regulatory Elements:\nTissue-Agnostic Model", "Transcriptome Prediction with 56 Regulatory Elements:\nTissue-Aware Model")
modAllData <- allR2;
mixtureAllDataFileName <- file.path(paths$scratch, 'mixtureAllData-ag-aw.Rda')


for (i in 1:ncol(allR2)) {
  label <- c('poor', 'well')
  expR2 <- colnames(allR2)[i]
  params <- bestBetaParameters[[expR2]]
  expTitle <- titles[i]
  tempData <- allR2[, expR2]
  tempData[tempData < 1E-10] <- 1E-10
  tempData[tempData > 1 - 1E-10] <- 1 - 1E-10
  
  if (bestK[i] > 1) {
    #for the first three models
    model <- bestBetaModels[[expR2]]
    tMember <- apply(model@Z, 1, which.max)
    low <- tempData < .5
    # swap poor/well labels
    poorComp <- paste(expR2, '.', 'poor', sep = "")
    wellComp <- paste(expR2, '.', 'well', sep = "")
    
    modAllData <- cbind(modAllData,  as.data.frame(matrix(rep(NA, 2*length(model@datavec)), ncol=2)))
    colnames(modAllData)[(ncol(modAllData) - 1):ncol(modAllData)] <- c(poorComp, wellComp)
    if (mean(tempData[tMember == 1]) > mean(tempData[tMember == 2])) {
      modAllData[[wellComp]][tMember == 1] <- tempData[tMember == 1 | !low]
      modAllData[[poorComp]][tMember == 2] <- tempData[tMember == 2 & low]
    } else {
      modAllData[[wellComp]][tMember == 2] <- tempData[tMember == 2 | !low]
      modAllData[[poorComp]][tMember == 1] <- tempData[tMember == 1 & low]
    }
  }
  
  p0 <- NULL
  x <- seq(0, 1, len = length(tempData))
  p0 <- ggplot(as.data.frame(tempData), aes(tempData)) +
    geom_histogram(
      aes(y = ..density..),
      color = "black",
      fill = "white",
      binwidth = .01
    ) +
    xlab(expression(R ^ 2)) + ylab('Density') + ggtitle(expTitle) + xlim(c(0, 1))
  for (j in 1:ncol(params)) {
    p0 <- p0 + stat_function(
      aes(x = x),
      fun = function(x, j)
      {
        params[1, j] * dbeta(x, shape1 = params[2, j], shape2 = params[3, j])
      },
      colour = myColors[j],
      n = 1000,
      size = 1.5,
      args = list(j = j)
    )
  }
  print(p0)  
}

save(modAllData, file = mixtureAllDataFileName)

```

### Tail Analysis
```{r tail , warning = FALSE}
for (i in 1:ncol(allR2)) {
  expR2 <- colnames(allR2)[i]
  f <- file.path(paths$scratch, paste(paste('tails_poor', expR2, sep = '_'),'.csv',sep = ''))
  g <- file.path(paths$scratch, paste(paste('tails_well', expR2, sep = '_'),'.csv',sep = ''))

  params <- bestBetaParameters[[expR2]]
  bMeans <- params[2,] / (params[2,] + params[3,])
  pIndex <- which.min(bMeans)
  wIndex <- which.max(bMeans)
  
  pThrsh <- qbeta(.05, shape1 = params[2, pIndex], shape2 = params[3, pIndex])
  wThrsh <- qbeta(.99, shape1 = params[2, wIndex], shape2 = params[3, wIndex])

  write.csv(rownames(modAllData)[allR2[,expR2] < pThrsh], file = f)
  write.csv(rownames(modAllData)[allR2[,expR2] > wThrsh], file = g)

}


```

## Tissue-Aware - Train/Test Split - Pick Number of Labels

```{r plot, warning = FALSE, echo=FALSE}
library(ggplot2)
nClusters <- c(1, 5, 10, 15, 20, 25, 30, 35)

plot.cv <- function(cvtrain, cvtest, ytext){
  yt <- colMeans(cvtest)
  et <- apply(cvtest, 2, sd)
  
  yn <- colMeans(cvtrain)
  en <- apply(cvtrain, 2, sd)
  
  toplot <- data.frame(yt = yt,et = et,yn = yn,en = en,nc = nClusters)
  
  ggplot(toplot) +
    geom_line(aes(x = nc, y = yt, col = "Test"), size = 1) +
    geom_point(aes(x = nc, y = yt, shape = 'a'), show.legend = FALSE) +
    geom_line(aes(x = nc, y = yn, col = "Train"), size = 1) +
    geom_point(aes(x = nc, y = yn, shape = 'c'), show.legend = FALSE) +
    geom_errorbar(aes(x = nc,ymin = yt - et,ymax = yt + et), width = 0.5) +
    geom_errorbar(aes(x = nc,ymin = yn - en,ymax = yn + en), width = 0.5) +
    theme_bw() + xlab("Number of Clusters") + ylab(ytext) +
    scale_color_manual(values = c("red", "green"), name = "")
}

load(file.path(paths$scratch, 'cvTables.Rda'))

plot.cv(cvRmseTrn, cvRmseTst, "RMSE")
plot.cv(cvCorrTrn, cvCorrTst, "Correlation Score")


toplotr2 <- data.frame(yt = colMeans(cvR2Trn),
                       et = apply(cvR2Trn, 2, sd),nc = nClusters)
ggplot(toplotr2) +
    geom_line(aes(x = nc, y = yt), size = 1.5) +
    geom_point(aes(x = nc, y = yt, shape = 'a'), show.legend = FALSE) +
    geom_errorbar(aes(x = nc,ymin = yt - et,ymax = yt + et), width = 0.5) +
    theme_bw() + xlab("Number of Clusters") + ylab("Average R2") 

rm(toplotr2, cvRmseTst, cvRmseTrn, cvCorrTrn, cvCorrTst, cvR2Trn)
rm(list=ls())
source("00-paths.R")
rm(tcgapaths, tfacpaths)
```


### TF and MiR Histograms for GTEx

```{r load medoids, warning = FALSE, echo=FALSE}
load(file.path(paths$scratch, paste('medoids_tf_all.Rda', sep = '')))
load(file.path(paths$scratch, paste('medoids_mir_all.Rda', sep = '')))
```

```{r gtf, warning = FALSE, echo=FALSE}
load(file.path(paths$scratch, "gtexTFMed.Rda"))
colnames(tfMedoid)[1] == colnames(gtexTFMed)[1]
```

#### TF Comparison
```{r hist comparison , warning = FALSE, echo=FALSE}
par(mfrow=c(1,2))
for(i in 1:28){
  hist(tfMedoid[,i], breaks=123, main="TF Medoid from TCGA", xlab=paste("TF", colnames(tfMedoid)[i]), freq=F )
  hist(gtexTFMed[,i], breaks=123, main="TF Medoid from GTEx", xlab=paste("TF", colnames(tfMedoid)[i]), freq=F)

}
```

#### Mir Comparison 

```{r miR comparison, warning = FALSE, echo=FALSE}
# Let's do the same for miRs. The point is that we can not eaily match all of the miRs so we just use the 18/28 that we can match directly in the first try. 

load(file.path(paths$clean, "gtexMir.Rda"))

converMirNames <- function (longMirs){
  shortMirs <- sapply(longMirs, function(x) substr(x, 9, nchar(x)))
  
  shortMirs <- sapply(shortMirs, function(x) {
    aa <- gregexpr(pattern = '-3p', x)
    index <- aa[[1]][[1]]
    return (ifelse(index != -1, substr(x, 1, index - 1), x))
  })
  
  shortMirs <- sapply(shortMirs, function(x) {
    aa <- gregexpr(pattern = '-5p', x)
    index <- aa[[1]][[1]]
    return (ifelse(index != -1, substr(x, 1, index - 1), x))
  })
  
  shortMirs <-
    sapply(shortMirs, function(x) paste("MIR", x, sep = ""))
  names(shortMirs) <- NULL
  shortMirs <- toupper(shortMirs)
  return(shortMirs)
}
shortMirs <- converMirNames(colnames(mirMedoid))

colnames(mirMedoid) <- shortMirs

easyMirs <- intersect(rownames(gtexMir), shortMirs)




gtexMirMed <- t(gtexMir[easyMirs, ])
dim(gtexMirMed)

par(mfrow=c(1,2))
for(i in 1:length(easyMirs)){
  curMir <- easyMirs[i]
  hist(mirMedoid[,curMir], breaks=123, main="miR Medoid from TCGA", xlab=paste("miR", curMir), freq=F )
  hist(gtexMirMed[,curMir], breaks=123, main="miR Medoid from GTEx", xlab=paste("miR", curMir), freq=F)
}
rm(list=ls())
gc()
source("00-paths.R")
```


### TF Only Result for GTEx
Here the goal is to ignore the miRs and also Train/Test. We assume that the correct number of parameters is 15 for clusters and we learn 15 clusters using only TFs. And we do not do train/test split, we use the whole TCGA data (for which we already know the thresher's output for TFs) once to learn 15 clusters and for each one of them a linear model. Then we test that linear model on GTEx.

```{r , warning = FALSE}
load(f <- file.path(paths$scratch,'tf_final_model_15.Rda'))
cat("TCGA-TF:")
cat(paste("\n    Averge R2:", round(mean(model$r2, na.rm = T),2)))
cat(paste("\n    Averge RMSE:", round(mean(model$rmse),2)))
cat(paste("\n    Averge Cor:", round((model$corr$acc),2)))

load(file.path(paths$scratch, 'tf_gtex_test_15.Rda'))
cat("\n GTEx:")
cat(paste("\n    Averge RMSE:", round(mean(gtexResults$rmse),2)))
cat(paste("\n    Averge Cor:", round((gtexResults$corr$acc),2)))
```

    
    

# Appendix

This analysis was performed using the following R packages.
```{r si}
sessionInfo()

```




