---
title: "Selecting TFs and Mirs from Xena's TCGA Data"
author: "Amir Asiaee"
date: '`r Sys.Date()`'
output:
  html_document:
    highlight: kate
    theme: yeti
    toc: yes
---

```{r globopt}
knitr:::opts_chunk$set(fig.path="../results/mirfigs/")
knitr::opts_chunk$set(fig.width=12) 
```

```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Getting Started

We run our feature selection pipeline (Thresehr and Reaper) on TCGA's miR and TF data onces for TCGA R^2 analysis. 

```{r globals}
rm(list=ls())
source("00-paths.R")

load(file.path(paths$clean, 'xena-tcga.Rda'))
dim(tcga$tf)
dim(tcga$mir)
```

# Thresher for MicroRNA
First, we investigate the number of zeros before aggregation of miRs and remove microRNAs that are zero in 80 percent of samples:

```{r , warning = FALSE}
# Getting a logical matrix indicating whether the value is 0
zero_logical_matrix <- tcga$mir == 0

# Calculating the percentage of zeros for each microRNA
zero_percentages <- rowMeans(zero_logical_matrix) * 100

nrow(tcga$mir) - sum(zero_percentages > 80)

selected_mirs <- tcga$mir[zero_percentages < 80, ]
dim(selected_mirs)

rm(zero_logical_matrix, zero_percentages)
```

Let's run Thresher on selected Mirs from TCGA:

```{r, warning = FALSE}
suppressMessages( library(Thresher) ) # brings along PCDimension automatically
f <- file.path(paths$scratch, "xena-tcga-thresh-mir.Rda")
if (file.exists(f)) {
  load(f)
} else {
  set.seed(47053)
  Thresh <- Thresher(t(selected_mirs), method="auer.gervini", 
                     scale=TRUE, agfun=agDimTwiceMean)
  Reap <- Reaper(Thresh, useLoadings=TRUE, cutoff = 0.35, metric=NULL, 
                 linkage="ward.D2", maxSampleGroups = 486)
  
  save(Thresh, Reap, file = f)
}

Thresh@pcdim  # Number of principal components
Reap@nGroups  # Number of clusters

groups <- predict(Reap@fit)
names(groups) <- rownames(selected_mirs)[Reap@keep]
table(groups)
NG <- Reap@nGroups
  
rm(f, Thresh, Reap)
```

## Select the Mir Medoids
We need to save the medoids: 
```{r , warning = FALSE}
f <- file.path(paths$scratch, "xena-tcga-mir-medoid.Rda")
if (file.exists(f)) {
  load(f)
} else {
  mirMedoid <- matrix(0, nrow=ncol(selected_mirs), ncol=NG)
  mirMedoidNames <- c()
  
  for (I in 1:NG) {
    clusterExp <- selected_mirs[names(groups)[groups == I],]
    meanZeroClusterExp <- sweep(clusterExp, 2, apply(clusterExp, 2, mean))
    distanceFromMean <- apply(meanZeroClusterExp, 1, function(x) {sqrt(sum(x*x))})
    medoidGene <- rownames(clusterExp)[which.min(distanceFromMean)]
    mirMedoid[, I] <- as.numeric(selected_mirs[medoidGene, ])
    mirMedoidNames <- c(mirMedoidNames, medoidGene)
  }
  colnames(mirMedoid) <- mirMedoidNames
  rownames(mirMedoid) <- colnames(selected_mirs)
  save(mirMedoid, file = f)
}

dim(mirMedoid)
print(colnames(mirMedoid))

rm(f, clusterExp, meanZeroClusterExp, distanceFromMean, medoidGene, 
   selected_mirs, mirMedoidNames, I, NG, groups)
```

# Thresher for TFs

Let's run Thresher on 478 TFs from TCGA:

```{r , warning = FALSE}
suppressMessages( library(Thresher) ) # brings along PCDimension automatically
f <- file.path(paths$scratch, "xena-tcga-thresh-tf.Rda")
if (file.exists(f)) {
  load(f)
} else {
  set.seed(47053)
  
  Thresh <- Thresher(t(tcga$tf), method="auer.gervini", 
                     scale=TRUE, agfun=agDimTwiceMean)
  Reap <- Reaper(Thresh, useLoadings=TRUE, cutoff = 0.35, metric=NULL, 
                 linkage="ward.D2", maxSampleGroups = 486)
  
  save(Thresh, Reap, file = f)
}

Thresh@pcdim  # Number of principal components
Reap@nGroups  # Number of clusters

groups <- predict(Reap@fit)
names(groups) <- rownames(tcga$tf)[Reap@keep]
table(groups)
NG <- Reap@nGroups
  
rm(f, Thresh, Reap)
```

## Select the TF Medoids
We need to save the medoids: 
```{r , warning = FALSE}
f <- file.path(paths$scratch, "xena-tcga-tf-medoid.Rda")
if (file.exists(f)) {
  load(f)
} else {
  tfMedoid <- matrix(0, ncol(tcga$tf), NG)
  tfMedoidNames <- c()
  
  for (I in 1:NG) {
    clusterExp <- tcga$tf[names(groups)[groups == I],]
    meanZeroClusterExp <- sweep(clusterExp, 2, apply(clusterExp, 2, mean))
    distanceFromMean <- apply(meanZeroClusterExp, 1, function(x) {sqrt(sum(x*x))})
    medoidGene <- rownames(clusterExp)[which.min(distanceFromMean)]
    tfMedoid[, I] <- as.numeric(tcga$tf[medoidGene,])
    tfMedoidNames <- c(tfMedoidNames, medoidGene)
  }
  colnames(tfMedoid) <- tfMedoidNames
  rownames(tfMedoid) <- colnames(tcga$tf)
  
  save(tfMedoid, file = f)
}

dim(tfMedoid)
print(colnames(tfMedoid))

rm(f, clusterExp, meanZeroClusterExp, distanceFromMean, medoidGene, 
   tfMedoidNames, I, NG, groups)
```


# Appendix

This analysis was performed using the following R packages.
```{r si}
sessionInfo()

```




