---
title: "Predicting the Whole Transcriptome"
author: "Amir Asiaee"
date: '`r Sys.Date()`'
output:
  html_document:
    highlight: kate
    theme: yeti
    toc: yes
---

```{r globopt}
knitr:::opts_chunk$set(fig.path="../results/mirfigs/")
knitr::opts_chunk$set(fig.width=12) 
```

```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Getting Started
Here, we continue to show that our selected features (TFs and miRs) are bilogically informative. We show that we can explain the variation of the whole transcriptome using only 47 measurements!

```{r globals}
rm(list=ls())
source("00-paths.R")

load(file.path(paths$clean, 'xena-tcga.Rda'))
load(file.path(paths$scratch, "xena-tcga-mir-medoid.Rda"))
load(file.path(paths$scratch, "xena-tcga-tf-medoid.Rda"))
mRNARest <- as.matrix(tcga$mrna)
# Removing constant genes to prevent Inf R^2
mRNAs_sd <- apply(mRNARest, 1, sd)
mRNARest <- mRNARest[mRNAs_sd != 0, ]

cancerType <- tcga$tissue

dim(mirMedoid)
dim(tfMedoid)
dim(mRNARest)

dev_mode <- T

rm(tcga, mRNAs_sd)
```

# R2 Joint Models
By joint model, we mean a model that uses both TFs and miRNAs 
```{r , warning = FALSE}
compute_and_save_outcome_variation <- function(f, outcome) {
  requireNamespace("ClassComparison", quietly = TRUE)  
  if (file.exists(f)) {
    load(f)
  } else {
    N <- ncol(outcome)
    dmean <- matrixMean(outcome)
    dvar <- as.vector(matrixVar(outcome, dmean))
    save(N, dvar, file = f)
  }
  return(list(N = N, dvar = dvar))
}

f <- file.path(paths$scratch, "xena-tcga-mRNA-outcome-constants.Rda")
constants <- compute_and_save_outcome_variation(f, mRNARest)

rm(f, const_genes, dmean, i, myZero, results)
```

```{r hist plot, warning = FALSE, echo=FALSE}
histsR2 <- function(residuals, saveName, constants, replace = F, thrsh = 0){
  f <- file.path(paths$scratch, paste(saveName, ".Rda", sep = ""))
  if (file.exists(f) && !replace) {
    load(f)
  } else {
    ssres <- apply(residuals^2, 1, sum)
    r2s <- 1 - ssres/(constants$dvar * (constants$N-1))
    save(r2s, file=f)
  }
  print(paste("Percentage of R2 more than the threshold ", thrsh, " is ", sum(r2s >thrsh) /length(r2s), sep = ""))
  par(mfrow=c(1,1))
  hist(r2s[r2s>thrsh], breaks=234, main="R2") 
  summary(r2s)
  return(r2s)
}

```

## Tissue-Agnostic Model

Here we regress each (of ~20,000) mRNA into miR and TF mediods. So we have 20,000 regression problems, 10,000 samples, and 47 features. 

```{r }
library(ClassComparison)

compute_agnostic_model <- function(f_path, g_path, X1, X2, Y, saveAs) {
  if (file.exists(f_path) && file.exists(g_path)) {
    load(f_path)
    load(g_path)
  } else {
    requireNamespace("ClassComparison", quietly = TRUE)  
    agnostic_mlm <- MultiLinearModel(Y ~ ., cbind(X1, X2), Y) #It has intercept internally
    agnostic_R <- Y - t(agnostic_mlm@predictions)
    save(agnostic_R, agnostic_mlm, file = g_path)
  }
  return(list(R = agnostic_R, mlm = agnostic_mlm))
}


f <- file.path(paths$scratch, "xena-tcga-agnostic-mlm.Rda")
g <- file.path(paths$scratch, "xena-tcga-agnostic-residuals.Rda")
agnostic <- compute_agnostic_model(f, g, tfMedoid, mirMedoid, mRNARest)
saveAs <- "xena-tcga-agnostic-R2"
agnostic_r2s <- histsR2(agnostic$R, saveAs, constants, replace = dev_mode)
```

We clean up in a separate block: 
```{r , warning = FALSE}
rm(agnostic, r2s, saveAs, ssres, f, g)
gc() 
```


## Tissue-Aware Model
An intelligent way of bringing in the tissue into the prediction is to have a separate prediction per tissue. We first mean zero the output (using all genes) and then fit a per tissue model (fit the the residuals). This is equivalent to fit a global linear model first and then fit the residuals with a linear model per tissue.

First let initialize:
```{r , warning = FALSE}
magicThresh <- 50  #We have 47 features, so at least we need 50 samples.
f <- file.path(paths$scratch, "xena-tcga-mean0mRNA.Rda")
if (file.exists(f)) {
  load(f)
} else {
  mean0mRNA <- sweep(mRNARest, 1, apply(mRNARest, 1, mean))
  save(mean0mRNA,  file = f)
}
rm(f)
```

Now, fit one linear model per tissue. For low sample tissues, the model is just an intercept. Some features are constants for some tissues, we remove them to avoid singularity:
```{r }
library(ClassComparison)
aware_residuals <- matrix(0, nrow = nrow(mean0mRNA), ncol = ncol(mean0mRNA))
f <- file.path(paths$scratch, "xena-tcga-tissue-aware-mlm.Rda")
g <- file.path(paths$scratch, "xena-tcga-tissue-aware-residuals.Rda")

if (file.exists(f)) {
  load(f)
  load(g)
} else {
  
  aware_mlm_list <- list()
  low_sample_tissue <- c() #what type has insufficient sample for MLM
  X <- cbind(tfMedoid, mirMedoid)
  variable_features_in_tissue <- matrix(T, length(unique(cancerType)), ncol(X))
  rownames(variable_features_in_tissue) <- unique(cancerType)
  
  for(type in unique(cancerType)){
    print(paste("Fitting", type))
    Ts <- (cancerType == type) #Tissue Selector
    if(sum(Ts) < magicThresh){ #Per-tissue average is our predictor   
      tissue_mrna_mean <- apply(mean0mRNA[,Ts], 1, mean)
      aware_residuals[,Ts] <- mean0mRNA[,Ts] - tissue_mrna_mean
      aware_mlm_list[[type]] <- tissue_mrna_mean #just an intercept
      low_sample_tissue <- c(low_sample_tissue, TRUE) 
    } else{
      newX <- X[Ts,]
      #Remove constant features to avoid singularity
      variable_features <- apply(newX, 2, min) != apply(newX, 2, max) #none degenerate (non-constant) features
      variable_features_in_tissue[type,] <- variable_features
      newX <- newX[, variable_features]
      aware_mlm <- MultiLinearModel(Y ~ ., newX, mean0mRNA[,Ts])
      aware_residuals[,Ts] <- mean0mRNA[,Ts] - t(aware_mlm@predictions)
      aware_mlm_list[[as.character(type)]] <- aware_mlm
      low_sample_tissue <- c(low_sample_tissue, FALSE) 
    }
    
  }
  # Checking low sample tissues:
  print(paste("Low sample cancer types", unique(cancerType)[low_sample_tissue]))
  # Sanity check
  table(cancerType)
  
  print(paste("Do we have any constant feature per tissue?", 
              ifelse(sum(apply(variable_features_in_tissue, 1, sum) != ncol(tfMedoid) + ncol(mirMedoid)) > 0, "Yes", "No")))
  
  save(aware_mlm_list,  file = f)
  save(aware_residuals, file = g)
}
rm(f, g)
gc()

saveAs <- "xena-tcga-aware-R2"
aware_r2s <- histsR2(aware_residuals, saveAs, constants, replace = dev_mode)
```

We clean up in a separate block. 
```{r , warning = FALSE}
rm(tfMedoid, mirMedoid, aware_mlm_list, aware_residuals, mean0mRNA, magicThresh, 
   variable_features, variable_features_in_tissue, low_sample_tissue, 
   dvar, newX, type, Ts, saveAs, X, tissue_mrna_mean, aware_mlm, cohorts, 
   mRNARest, N, cancerType, tcga)
gc()
```


# BetaMixture Model


```{r bestk, warning = FALSE, , echo=FALSE}
library(BetaModels)

allR2 <- data.frame(agnostic = agnostic_r2s, aware = aware_r2s)

#Fitting the best K once with more precision:
library(fitdistrplus)
bestK <- c(2, 2) #manually entered 
bestBetaParameters <- bestBetaModels <- list()


bestModelsFileName <- file.path(paths$scratch, 'bestBetaModels-final-ag-aw-tpm.Rda')
bestParamsFileName <- file.path(paths$scratch, 'bestBetaParams-final-ag-aw-tpm.Rda')
if (!file.exists(bestParamsFileName)) {
  for (i in 1:ncol(allR2)) {
    expR2 <- colnames(allR2)[i]
    tempData <- allR2[, expR2]
    tempData[tempData < 1E-10] <- 1E-10
    tempData[tempData > 1 - 1E-10] <- 1 - 1E-10
    myK <- bestK[i]
    
    if (myK == 1) {
      output <- fitdist(tempData, "beta", "mle")
      params <- matrix(c(1, output$estimate), ncol = 1)
    } else {
      output <-
        BetaMixture(tempData,
                    K = myK,
                    forever = 100,
                    epsilon = 0.001)
      params <- rbind(output@phi, output@mle)
    }
    bestBetaModels[[expR2]] <- output
    rownames(params) <- c("pi", "alpha", "beta")
    bestBetaParameters[[expR2]] <- params
  }
  save(bestBetaParameters, file = bestParamsFileName)
  save(bestBetaModels, file = bestModelsFileName)
  
} else {
  load(bestParamsFileName)
  load(bestModelsFileName)
}
```

```{r , warning = FALSE, echo=FALSE}

library(ggplot2)
library(Polychrome)
myColors <- light.colors(5)
titles <-
  c("Transcriptome Prediction with 47 Regulatory Elements:\nTissue-Agnostic Model", "Transcriptome Prediction with 47 Regulatory Elements:\nTissue-Aware Model")
modAllData <- allR2;
mixtureAllDataFileName <- file.path(paths$scratch, 'mixtureAllData-ag-aw-tpm.Rda')


for (i in 1:ncol(allR2)) {
  label <- c('poor', 'well')
  expR2 <- colnames(allR2)[i]
  params <- bestBetaParameters[[expR2]]
  expTitle <- titles[i]
  tempData <- allR2[, expR2]
  tempData[tempData < 1E-10] <- 1E-10
  tempData[tempData > 1 - 1E-10] <- 1 - 1E-10
  
  if (bestK[i] > 1) {
    #for the first three models
    model <- bestBetaModels[[expR2]]
    tMember <- apply(model@Z, 1, which.max)
    low <- tempData < .5
    # swap poor/well labels
    poorComp <- paste(expR2, '.', 'poor', sep = "")
    wellComp <- paste(expR2, '.', 'well', sep = "")
    
    modAllData <- cbind(modAllData,  as.data.frame(matrix(rep(NA, 2*length(model@datavec)), ncol=2)))
    colnames(modAllData)[(ncol(modAllData) - 1):ncol(modAllData)] <- c(poorComp, wellComp)
    if (mean(tempData[tMember == 1]) > mean(tempData[tMember == 2])) {
      modAllData[[wellComp]][tMember == 1] <- tempData[tMember == 1 | !low]
      modAllData[[poorComp]][tMember == 2] <- tempData[tMember == 2 & low]
    } else {
      modAllData[[wellComp]][tMember == 2] <- tempData[tMember == 2 | !low]
      modAllData[[poorComp]][tMember == 1] <- tempData[tMember == 1 & low]
    }
  }
  
  p0 <- NULL
  x <- seq(0, 1, len = length(tempData))
  p0 <- ggplot(as.data.frame(tempData), aes(tempData)) +
    geom_histogram(
      aes(y = ..density..),
      color = "black",
      fill = "white",
      binwidth = .01
    ) +
    xlab(expression(R ^ 2)) + ylab('Density') + ggtitle(expTitle) + xlim(c(0, 1))
  for (j in 1:ncol(params)) {
    p0 <- p0 + stat_function(
      aes(x = x),
      fun = function(x, j)
      {
        params[1, j] * dbeta(x, shape1 = params[2, j], shape2 = params[3, j])
      },
      colour = myColors[j],
      n = 1000,
      size = 1.5,
      args = list(j = j)
    )
  }
  print(p0)  
}

save(modAllData, file = mixtureAllDataFileName)

```

# Appendix

This analysis was performed using the following R packages.
```{r si}
sessionInfo()
```




