
---
title: "Generalizability and Transportability of Transcriptome Prediction"
author: "Amir Asiaee"
date: '`r Sys.Date()`'
output:
  html_document:
    highlight: kate
    theme: yeti
    toc: yes
---

```{r globopt, echo=FALSE}
knitr::opts_chunk$set(fig.width=12) 
```

```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Getting Started

One reason that makes the transportability a failure, is the inconsistency between the distributions of the matched miRs. So the part of the model that uses miRs will even hurt the prediction. To mitigate this issue, we won't consider miRs here and train a model solely on TFs. Here we will not redo the cross-validation for determining the number of tissues and use 15 tissues to train the model. This is can only hurt our model because target datsets are separate, unseen test data sets. 


```{r globals}
rm(list=ls())
source("00-paths.R")
library(ClassComparison)

ls()
```

# Training 

Note that since miR and TF features are selected separately, we do not need to redo feature selections here. We just remove miRs from consideration. 

```{r , warning = FALSE}
load(file.path(paths$clean, 'xena-tcga.Rda'))
load(file.path(paths$scratch, paste('medoids_tf_all_tpm.Rda', sep = '')))
dim(tfMedoid)
dim(tcga$mrna)
mRNARest <- as.matrix(tcga$mrna)
```

## Regression

```{r helpers, warning = FALSE, echo=FALSE}
computeCor <- function(res, Y, ns=10^4){
  res <- as.matrix(res)
  Y <- as.matrix(Y)
  Yhat <- Y - res
  M <- nrow(Y)
  Ours <- sapply(1:M, function(i) cor(Y[i,], Yhat[i,]))
  
  Yind <- sample(1:M, ns, replace = T)
  YhatInd <- sample(1:M, ns, replace = T)
  ind <- cbind(Yind, YhatInd)
  ind <- ind[Yind != YhatInd, ]
  Null <- sapply(1:nrow(ind), function(j) cor(Y[ind[j,1],], Yhat[ind[j,2],]))
  z <- quantile(Null, .95, na.rm = T)
  acc <- sum(Ours > z, na.rm = T) / M
  
  # Create separate data frames
  df_ours <- data.frame(value = Ours, category = "Ours")
  df_null <- data.frame(value = Null, category = "Null")

  # Combine data frames
  df <- rbind(df_ours, df_null)

  # Plot the histograms
  library(ggplot2)
  p <- ggplot(df, aes(x = value, fill = category)) +
    geom_histogram(aes(y = ..density..), position = "identity", alpha = 0.5, bins = 30) +
    labs(title = "Histogram of Ours and Null",
         x = "Value",
         y = "Frequency") +
    scale_fill_manual(values = c("blue", "red")) +
    theme_minimal()

  print(p)
  
  return(list(ours=Ours, null=Null, thresh=z, acc=acc))
}

computeR2 <- function(res, Y){
  res <- as.matrix(res)
  Y <- as.matrix(Y)
  M <- ncol(Y)
  dmean <- matrixMean(Y)
  dvar <- as.vector(matrixVar(Y, dmean))
  ssres <- apply(res^2, 1, sum)
  r2 <- rep(0, M)
  # varInd <- dvar != 0
  # r2[varInd] <- 1 - ssres[varInd] / (dvar[varInd] * (M - 1))
  r2 <- 1 - ssres / (dvar * (M - 1))
  return(r2)
}

compute_cor_acc <- function(residuals, y, Null, thresh){
  set.seed(12345)
  yhat <- y - residuals
  M <- nrow(y)
  Ours <- sapply(1:M, function(i) cor(y[i,], yhat[i,]))
  cat("\n Ours info:\n")
  cat(summary(Ours))
  
  acc <- sum(Ours > thresh, na.rm = T) / M

  # Create separate data frames
  df_ours <- data.frame(value = Ours, category = "Ours")
  df_null <- data.frame(value = Null, category = "Null")

  # Combine data frames
  df <- rbind(df_ours, df_null)

  # Plot the histograms
  library(ggplot2)
  p <- ggplot(df, aes(x = value, fill = category)) +
    geom_histogram(aes(y = ..density..), position = "identity", alpha = 0.5, bins = 30) +
    labs(title = "Histogram of Ours and Null",
         x = "Value",
         y = "Frequency") +
    scale_fill_manual(values = c("blue", "red")) +
    theme_minimal()

  print(p)

  # return(list(ours=Ours, null=Null, acc=acc))
  return(acc)
}

```

Training on TCGA data using only TF:
```{r , warning = FALSE}

nc <- 15
f <- file.path(paths$scratch, paste('final_model_tf_only', nc, '.Rda', sep = ''))
if (!file.exists(f)) {
  set.seed(1234)
  Yavg <- as.matrix(apply(mRNARest, 1, mean))
  Y <- as.matrix(sweep(mRNARest, 1, Yavg))
  X <- as.matrix(tfMedoid)
  
  coeffMatList <- list()
  N <- nrow(X); M <- nrow(Y); P <- ncol(X);
  res <- matrix(0, nrow = M, ncol = N)
  
  cat(paste("\n  Learning a model with ", nc, " clusters:"))
  cat("\n    1) Clustering")
  myClust <- kmeans(X, nc, iter.max = 100, nstart = 100)
  assignment <- myClust$cluster

  cat("\n    2) Regression")
  cat("\n      Fitting OLS to clusters 1")
  
  for (cluster in 1:nc) {
    if(cluster %% 5 == 0) cat(paste(cluster)) else cat('.')
    coeffMat <- matrix(0, nrow = P + 1, ncol = M, 
                       dimnames = list(c('(Intercept)',
                                         gsub('\\-', '.', colnames(X))), 
                                         rownames(Y)))
    
    S <- (assignment == cluster) #Cluster Selector
    varCov <- apply(X[S,], 2, var) != 0
    Xs <- X[S, varCov]
     # Create a mapping from variable genes to their positions in the original data
    mapping <- which(varCov)
    Ys <- Y[, S]
    
    if (nrow(Xs) < ncol(Xs)) {
      print("Not enough samples, using the means.")
      coeffMat[1,] <- apply(Ys, 1, mean)#YcAvg 
      res[, S] <- sweep(Ys, 1, coeffMat[1,])#mean0Yc 
    } else{
      mlm <- MultiLinearModel(Y ~ ., Xs, Ys)
      # Update the coeffMat only for variable genes, leaving others as zero
      coeffMat[1,] <- mlm@coefficients[1, ]  # +1 to account for the '(Intercept)' row
      coeffMat[mapping + 1,] <- mlm@coefficients[-1, ]  # +1 to account for the '(Intercept)' row
      res[, S] <- Ys - t(mlm@predictions)
    }
    coeffMatList[[cluster]] <- coeffMat
  } 
  
  r2 <- computeR2(res, Y)
  cat(paste("\n      Average R2:", round(mean(r2, na.rm =T), 2)))
  rmse <- sqrt((1/N) * apply(res^2, 1, sum))
  cat(paste("\n      Averge RMSE:", round(mean(rmse),2)))
  corr <- computeCor(res, Y)
  cat(paste("\n      Averge Cor:", round(mean(corr$acc),2)))
  
  model <- list(coeffMatList=coeffMatList, nc=nc,
                  res=res, r2=r2, rmse=rmse, corr=corr,
                  centers=myClust$centers, alpha=Yavg)

  save(model, file = f)
} else {
  load(f)
}

hist(model$r2, breaks=123, main="R2", xlab=paste("Average R2 =", round(mean(model$r2, na.rm =T), 2)))
print(paste("\n      Average R2:", round(mean(model$r2, na.rm =T), 2)))
print(paste("\n      Average RMSE:", round(mean(model$rmse, na.rm =T), 2)))
print(paste("\n      Average Cor Score:", round(model$corr$acc, 2)))


rm(coeffMat, coeffMatList, mlm, res, r2, rmse, corr, Yavg, myClust, Y, Ys, X, Xs)
```


# Testing
First let's select the corresponding submatrix of data for the three test datasets. 
```{r , warning = FALSE}
load(file.path(paths$clean, 'ccle.Rda'))
```

## CCLE

### Preparing Input: X

```{r , warning = FALSE}
tf_name_cleaner <- function(name){(sub("\\..*$", "", name))}

find_tcga_medoid_in_ccle <- function(tcga_medoid_names, ccle_feature_names, name_cleaner, d2MeanList, tcga_data, ccle_data){
  
  matched_names <- vector("character", length(tcga_medoid_names))
  for (i in seq_along(tcga_medoid_names)) {
    name <- tcga_medoid_names[i]
    d2Mean <- d2MeanList[[i]]
    d2Mean_sorted <- sort(d2Mean)
    trial <- 1
    
    while(TRUE) {
      trial <- trial + 1
      # hist(as.numeric(tcga_data[name,]), main=paste('tcga', name), breaks = 123)
          
      if (name %in% ccle_feature_names) {
        matched_names[i] <- name
        break
      }
      
      name_clean <- name_cleaner(name)
      # Sometimes mirA-3p has matched to mirA and now mirA-5p is considered
      # We want to avoid mirA-5p going to mirA again, so we skip it.
      if (name_clean %in% ccle_feature_names && !(name_clean %in% matched_names)) {
        matched_names[i] <- name_clean
        break
      }
      next_smallest_value <- d2Mean_sorted[trial]
      name <- names(d2Mean)[which(d2Mean == next_smallest_value)]
    }
    # hist(as.numeric(ccle_data[matched_names[i], ]), main=paste('ccle', matched_names[i]), breaks = 123)
    # print(paste("Converted", name, "in trial", trial - 1, "out of", length(d2Mean), "to", matched_names[i]))
  }
  return(matched_names)
  
}

load(file.path(paths$scratch, paste('d2MeanList_tf_all.Rda', sep = '')))
cleaned_ccle_tf_names <- sapply(rownames(ccle$tf), tf_name_cleaner)
print("Converting TF names:")
rownames(ccle$tf) <- cleaned_ccle_tf_names
ccle_tf_medoid_names <- find_tcga_medoid_in_ccle(colnames(tfMedoid), cleaned_ccle_tf_names, tf_name_cleaner, d2MeanList, tcga$tf, ccle$tf)

ccle_tf_medoid <- t(as.matrix(ccle$tf[ccle_tf_medoid_names, ]))
```

### Preparing Output: Y
```{r , warning = FALSE}
tcga_mrna_names <- sapply(rownames(tcga$mrna), tf_name_cleaner)
ccle_mrna_names <- sapply(rownames(ccle$mrna), tf_name_cleaner)

sum(ccle_mrna_names %in% tcga_mrna_names)
length(ccle_mrna_names)
```
It seems we can predict everything. 

Getting same names and order for outcomes:

```{r const computation, warning = FALSE}
library(ClassComparison)

f <- file.path(paths$scratch, "Y_ccle.Rda")
if (file.exists(f)) {
  load(f)
} else {
  # common ordering of names for future
  Ynames <- intersect(tcga_mrna_names, ccle_mrna_names)
  Y_ccle <- as.matrix(ccle$mrna[Ynames, ])
  save(Y_ccle, Ynames, file = f)
}

length(Ynames)
dim(Y_ccle)
```

### Prediction

```{r , warning = FALSE}
f <- file.path(paths$scratch, paste('ccle_test_tf_only', model$nc, '.Rda', sep = ''))

if(!file.exists(f)){
  cleaned_mrna_names <- sapply(rownames(model$alpha), tf_name_cleaner)
  rownames(model$alpha) <- cleaned_mrna_names
  Y <- sweep(Y_ccle, 1, model$alpha[Ynames,]) #removing global model intercept
  # Y <- sweep(Y_ccle, 1, apply(Y_ccle, 1, mean)) #removing global model intercept
  X <- ccle_tf_medoid
  
  centers <- model$centers
  
  cat(paste("\n  Testing with ", nrow(centers), " clusters:"))
  M <- nrow(Y); N <- ncol(Y); #nrow(X) == N
  D <- matrix(0, nrow = N, ncol = nrow(centers))
  for(i in 1:nrow(centers)){
    center <- centers[i, ]
    D[,i] <- apply(X, 1, function(x) sum((x - center)^2))
  }
  C <- apply(D, 1, which.min)
  
  res <- matrix(0, nrow = M, ncol = N)
  aX <- cbind(rep(1, nrow(X)), X) #augmented X
  for(i in 1:nrow(centers)){
    print(paste("Cluster", i, "has size", sum(C==i)))
    if(sum(C==i) == 0) next;
    Beta <- model$coeffMatList[[i]]
    colnames(Beta) <- cleaned_mrna_names
    res[, C==i] <- Y[,C==i] - t(aX[C==i, ] %*% Beta[,Ynames])
  }
  rmse <- sqrt((1/N) * apply(res ^ 2, 1, sum))
  cat(paste("\n    Averge RMSE:", round(mean(rmse),2)))
  corr <- computeCor(res, Y_ccle)
  cat(paste("\n    Averge Cor:", round(corr$acc,2)))
  ccle_results <- list(res=res, rmse=rmse, corr=corr)
  save(ccle_results, file = f)
} else{
  load(f)
}

cat(paste("\n    Averge RMSE:", round(mean(ccle_results$rmse),2)))
cat(paste("\n    Averge Cor:", round((ccle_results$corr$acc),2)))
```

## GTEx and TARGET

### Prediction GTEx

```{r , warning = FALSE}
f <- file.path(paths$scratch, paste('gtex_test_tf_only', model$nc, '.Rda', sep = ''))
load(file.path(paths$clean, 'xena-gtex.Rda'))


if(!file.exists(f)){
  Y <- sweep(as.matrix(gtex$mrna), 1, model$alpha) #removing global model intercept
  X <- t(as.matrix(gtex$tf))
  X <- X[, colnames(tfMedoid)]
  
  centers <- model$centers
  
  cat(paste("\n  Testing with ", nrow(centers), " clusters:"))
  M <- nrow(Y); N <- ncol(Y); #nrow(X) == N
  D <- matrix(0, nrow = N, ncol = nrow(centers))
  for(i in 1:nrow(centers)){
    center <- centers[i, ]
    D[,i] <- apply(X, 1, function(x) sum((x - center)^2))
  }
  C <- apply(D, 1, which.min)
  
  res <- matrix(0, nrow = M, ncol = N)
  aX <- cbind(rep(1, nrow(X)), X) #augmented X
  for(i in 1:nrow(centers)){
    print(paste("Cluster", i, "has size", sum(C==i)))
    if(sum(C==i) == 0) next;
    Beta <- model$coeffMatList[[i]]
    res[, C==i] <- Y[,C==i] - t(aX[C==i, ] %*% Beta)
  }
  rmse <- sqrt((1/N) * apply(res ^ 2, 1, sum))
  cat(paste("\n    Averge RMSE:", round(mean(rmse),2)))
  # corr <- computeCor(res, Y_ccle)
  corr <- compute_cor_acc(res, Y, model$corr$null, model$corr$thresh)
  cat(paste("\n    Averge Cor:", round(corr,2)))
  gtex_results <- list(res=res, rmse=rmse, corr=corr)
  save(gtex_results, file = f)
} else{
  load(f)
}

cat(paste("\n    Averge RMSE:", round(mean(gtex_results$rmse),2)))
cat(paste("\n    Averge Cor:", round((gtex_results$corr),2)))
```

### Prediction TARGET

```{r , warning = FALSE}
f <- file.path(paths$scratch, paste('target_test_tf_only', model$nc, '.Rda', sep = ''))
load(file.path(paths$clean, 'xena-target.Rda'))


if(!file.exists(f)){
  Y <- sweep(as.matrix(target$mrna), 1, model$alpha) #removing global model intercept
  X <- t(as.matrix(target$tf))
  X <- X[, colnames(tfMedoid)]
  
  centers <- model$centers
  
  cat(paste("\n  Testing with ", nrow(centers), " clusters:"))
  M <- nrow(Y); N <- ncol(Y); #nrow(X) == N
  D <- matrix(0, nrow = N, ncol = nrow(centers))
  for(i in 1:nrow(centers)){
    center <- centers[i, ]
    D[,i] <- apply(X, 1, function(x) sum((x - center)^2))
  }
  C <- apply(D, 1, which.min)
  
  res <- matrix(0, nrow = M, ncol = N)
  aX <- cbind(rep(1, nrow(X)), X) #augmented X
  for(i in 1:nrow(centers)){
    print(paste("Cluster", i, "has size", sum(C==i)))
    if(sum(C==i) == 0) next;
    Beta <- model$coeffMatList[[i]]
    res[, C==i] <- Y[,C==i] - t(aX[C==i, ] %*% Beta)
  }
  rmse <- sqrt((1/N) * apply(res ^ 2, 1, sum))
  cat(paste("\n    Averge RMSE:", round(mean(rmse),2)))
  # corr <- computeCor(res, Y_ccle)
  corr <- compute_cor_acc(res, Y, model$corr$null, model$corr$thresh)
  cat(paste("\n    Averge Cor:", round(corr,2)))
  target_results <- list(res=res, rmse=rmse, corr=corr)
  save(target_results, file = f)
} else{
  load(f)
}

cat(paste("\n    Averge RMSE:", round(mean(target_results$rmse),2)))
cat(paste("\n    Averge Cor:", round((target_results$corr),2)))
```



# Appendix

This analysis was performed using the following R packages.
```{r si}
sessionInfo()

```



