
---
title: "Tissue Biomarker Prediction"
author: "Amir Asiaee"
date: '`r Sys.Date()`'
output:
  html_document:
    highlight: kate
    theme: yeti
    toc: yes
---

```{r globopt, echo=FALSE}
knitr::opts_chunk$set(fig.width=12) 
```

```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Getting Started

The goal here is to determine the genes that are biomarkers of tissues from external resources and show that we are predicting those genes specifically very well. 

```{r globals}
rm(list=ls())
source("00-paths.R")
ls()
```

# External Data Source
Here we used "The Human Protein Atlas" to come up with the tissue markers. Using the following query we extracted the TSV file from their website:
`tissue_category_rna:any; Tissue enriched AND sort_by:tissue specific score.` The `Tissue enriched` flag tells to select genes that are enriched for any tissue. 

Here, we load the downloaded data:

```{r , warning = FALSE}
library(readr)

hpa <- read_delim(file.path(paths$raw, "tissue_category_rna_any_Tissue.tsv"), 
                       delim = "\t", col_names = TRUE, show_col_types = FALSE)
ncol(hpa)
colnames(hpa)[1:20]

head(hpa)[18:19]
hpa_data <- hpa[,c(1, 3, 18:19)]
```
The file is sorted based on the tissue specificity score and we also need to extract the tissue name from the 19th column which has both the tissue name and normalized TPM (for protein level expression)

```{r , warning = FALSE}
library(dplyr)
library(tidyr)

hpa_parsed <- hpa_data %>%
  separate(`RNA tissue specific nTPM`,
           into = c("TissueName", "ExpressionValue"),
           sep = ": ", 
           extra = "merge", 
           fill = "right") %>% 
  mutate(ExpressionValue = as.numeric(ExpressionValue))

head(hpa_parsed)
```

Now we need to match the tissue name to that of TCGA's samples. 
```{r , warning = FALSE}
unique(hpa_parsed$TissueName)
```
```{r , warning = FALSE}
load(file.path(paths$clean, 'xena-tcga.Rda'))
cancerType <- as.factor(tcga$tissue)
unique(cancerType)
rm(tcga)
gc()
```
Now, afte rmanual inspection, map the tissues that are similar 
```{r , warning = FALSE}
mappingDF <- tibble(
  TissueName = c("breast", "liver", "pancreas",
                 "thyroid gland", "prostate", "ovary", "endometrium 1",
                 "urinary bladder", "esophagus", "stomach 1"),
  TCGA_Code  = c("BRCA",   "LIHC",  "PAAD",
                 "THCA",   "PRAD", "OV",    "UCEC",
                 "BLCA",   "ESCA", "STAD")
)

hpa_filtered <- hpa_parsed %>%
  inner_join(mappingDF, by = "TissueName")


nTPM_cutoff <- 100
hpa_final <- hpa_filtered %>%
  filter(ExpressionValue >= nTPM_cutoff)

head(hpa_final)
rm(hpa, hpa_data, hpa_filtered, hpa_parsed, mappingDF)
```

# Loading TCGA Prediction Results

```{r , warning = FALSE}
load(file.path(paths$scratch, 'xena-tcga-agnostic-R2.Rda'))
r2_tcga_agnostic <- r2s
rm(r2s)

load(file.path(paths$scratch, 'xena-tcga-aware-R2.Rda'))
r2_tcga_aware <- r2s
rm(r2s)

# For some reason the gene ids are missing in the aware model
names(r2_tcga_aware) <- names(r2_tcga_agnostic)

head(r2_tcga_agnostic)
head(r2_tcga_aware)
```
Let's unify the Ensembl id naming:

```{r , warning = FALSE}
remove_version <- function(x) sub("\\..*$", "", x)

# Create data.frames with columns "ensembl_id" and "r2"
r2_agnostic_df <- data.frame(
  ensembl_id = remove_version(names(r2_tcga_agnostic)),
  r2_agnostic = as.numeric(r2_tcga_agnostic),
  stringsAsFactors = FALSE
)

r2_aware_df <- data.frame(
  ensembl_id = remove_version(names(r2_tcga_aware)),
  r2_aware = as.numeric(r2_tcga_aware),
  stringsAsFactors = FALSE
)

```

We'll do an inner_join on the "Ensembl" field from hpa_final and the newly created "ensembl_id" columns. That way, we only keep genes that appear in both data sets.


```{r , warning = FALSE}
hpa_final_merged <- hpa_final %>%
  rename(ensembl_id = Ensembl) %>%
  inner_join(r2_agnostic_df, by = "ensembl_id") %>%
  inner_join(r2_aware_df, by = "ensembl_id")

head(hpa_final_merged)
```

Now, plotting:

```{r , warning = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr) 

hpa_long <- hpa_final_merged %>%
  select(ensembl_id, Gene, TissueName, TCGA_Code, r2_agnostic, r2_aware) %>%
  gather(key = "Model", value = "R2", r2_agnostic, r2_aware) %>%
  mutate(Model = recode(Model, 
                        "r2_agnostic" = "Tissue-Agnostic",
                        "r2_aware" = "Tissue-Aware"))

p_global <- ggplot(hpa_long, aes(x = Model, y = R2)) +
  geom_violin(trim = FALSE, fill = "lightblue") +
  geom_boxplot(width = 0.1, outlier.shape = NA) +
  theme(plot.title = element_text(size = 10)) +
  theme_minimal(base_size = 14) +
  labs(title = "Comparison of R² for Tissue-Enriched Marker Genes",
       x = "Model",
       y = expression(R^2)) +
  # Add Wilcoxon signed-rank test annotation
  stat_compare_means(method = "wilcox.test", paired = TRUE, 
                     label = "p.signif", 
                     label.y = 1.0)  # adjust label.y as needed

p_global

fix_tissue_name <- function(x) {
  x <- recode(x, 
              "stomach 1" = "Stomach",
              "urinary bladder" = "Bladder",
              .default = x)
  # Capitalize first letter if you'd like (for any not recoded above):
  substring(x, 1, 1) <- toupper(substring(x, 1, 1))
  x
}
  

hpa_long <- hpa_long %>%
  mutate(TissueNameClean = fix_tissue_name(TissueName))

library(tibble)
cohorts_joinable <- cohorts %>%
  rownames_to_column("TCGA_Code")  # or rename col if needed

hpa_long_colored <- hpa_long %>%
  left_join(select(cohorts_joinable, TCGA_Code, color),
            by = "TCGA_Code")


p_facet <- ggplot(hpa_long_colored, aes(x = Model, y = R2, fill = color)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white", alpha = 0.4) +
  scale_fill_identity() +
  facet_wrap(~ TissueNameClean, scales = "free_y") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Comparison of R² for Tissue-Enriched Marker Genes (Per-Tissue)",
    x = "Model",
    y = expression(R^2)
  )

p_facet


wilcox_test_res <- wilcox.test(
  hpa_final_merged$r2_agnostic,
  hpa_final_merged$r2_aware,
  paired = TRUE
)

library(patchwork)
p_global_no_title <- p_global + ggtitle(NULL)
p_facet_no_title  <- p_facet + ggtitle(NULL)

p_combined <- p_global_no_title + p_facet_no_title +
  plot_layout(ncol = 2, widths = c(1, 3)) +  # adjust widths as needed
  plot_annotation(title = "Global and Per-Tissue Comparisons of R² for Tissue-Enriched Marker Genes")

p_combined

ggsave(file.path(paths$scratch, "p_combined.png"), p_combined, width = 12, height = 6, dpi = 300)
```

# Appendix

This analysis was performed using the following R packages.
```{r si}
sessionInfo()

```



